Extracting Function Signatures from Program Traces

### 2020-05-14 paper/main.tex
reorganizing

@@

\subsection{contractr}
\label{sec:contractr}

@@

%
%
\subsubsection{Overview}
The goal of our analysis is to discover argument types for functions,
builtins, and specials.  Naively intercepting function calls and recording
argument types is incorrect, as recall that R wraps arguments in promises
(Section \AT{what section}).  To circumvent this obstacle, we instead
intercept the forcing of promises: when a promise is forced, we identify the
surrounding call, collect the type of the value, and add it to the call
trace associated with the current function call.  Once function execution
completes, we collect information about the return value, add it to the call
trace, and save the completed call trace to be written out later.  \AT{This
  paragraph overlaps with the initial discussion of \contractr. We need to
  make sure that we don't inadvertently repeat ourselves.}

%
%
\subsubsection{Type Information}
Once a value is obtained, either on function return or when an argument
promise is forced, we probe the value for the following information:

\item First, we wanted to ensure that \typetracer captured type for
  unevaluated arguments.  Not every argument of a function needs to be used
  during every function execution, and the approach we discussed above
  relies on argument promises being forced (and, consequently, on arguments
  being used).  There are a few reasons for this.

One possibility is that the argument was simply unused given the control
flow of this particular execution (e.g. if the argument is only needed in
one branch of an \code{if}).  This mainly occurs when arguments have default
values for some parameters that are not critical to every execution oft he
function.  In these cases, we still want to collect the type of the default
value.

Another possibility is that a value is used {\it without being forced}, as
is the case when values are ``metaprogrammed'': \AT{describe, code example.}

\item Second, we wanted to account for formal arguments (arguments which are
  named in the function definition) that were {\it not} passed to the
  function.  Put simply, a function was not called with all of its
  arguments, and no default value was specified for missing arguments.  In
  this case, we record a ``missing'' type for the argument.

There are two obvious ways to deal with missing arguments: type them as bot,
or as top.  As we are performing a dynamic analysis, based on package test
code, we conservatively type these arguments as top, or \code{any}.  It is
impossible to know dynamically what all possible function behaviours are,
\AT{I'm still not convinced about this.}

\item Third, we had to contend with R's C implementation sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  These longjumps occur quite often, particularly when S3
  dispatch occurs, as the R implementation optimizes returns from dispatched
  functions by initiating a longjump.

We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
\texttt{special\_entry},\texttt{special\_exit},
\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
callbacks fire when closures, builtin, and special functions are entered and
exited, and before and after the forcing of promises.  The function-related
callbacks are used mainly for bookkeeping: the analysis is notified that a
closure, builtin, or special has been entered/exited by pushing/popping the
call onto a stack.  The calls themselves store a {\it call trace} object,
where we store the type information that we collect, populated with initial
guesses of argument types (as discussed in
Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
is recorded on the {\tt \_entry} variants: \AT{recall} that there are
multiple ways that R performs dynamic dispatch, based on the class of
function arguments.  We collect dispatch information in order to analyze
dispatch usage patterns.

Once types have been simplified in this way, for each function we move to
consolidate its multiple call traces into a single function type.  We do
this by combining the observed types of each argument into one type
(possibly a union, mind), according to subtyping rules that are part of our
design.  The precise subtyping rules we used will be discussed further in
Section~\ref{sec:typesystemdesign}.

While our analysis collects more detailed information than necessary for
type generation, the information proved useful for building up a usage
profile of the language. Through it, we were able to determine the best
translation rules (i.e., determining the type system we are translating to)
through an iterative process, involving analyzing usage patterns that arose
in the data and making design decisions based on them, revisiting the data
as necessary.  In the following section, we will discuss these decisions in
more detail, highlighting the reasoning behind our type system design.

@@

@@

@@

%
%
\subsubsection{Use of \code{any}}

>>>>

\AT{Leaving this one for Aviral.}

@@

In this section, we will describe our approach to data collection and analysis through \typetracer, a dynamic analysis built in the \rdt framework.
We will then describe the process by which we distill the rich information we collect with \typetracer to more digestible ``types'', though we discuss the precise details of our type annotation language in Section~\ref{sec:types}.
Finally, we discuss \contractr, a package built in R which supports our type language and can generate and check contracts for function types.

%
%
\subsubsection{Overview}
The goal of our analysis is to discover argument types for functions,
builtins, and specials.  Naively intercepting function calls and recording
argument types is incorrect, as recall that R wraps arguments in promises
(Section~\ref{sec:R}).  To circumvent this obstacle, we instead
intercept the forcing of promises: when a promise is forced, we identify the
surrounding call, collect the type of the value, and add it to the call
trace associated with the current function call.  Once function execution
completes, we collect information about the return value, add it to the call
trace, and save the completed call trace to be written out later.
%
%
\subsubsection{Type Information}
Once a value is obtained, either on function return or when an argument
promise is forced, we probe the value for the following information:

\item First, we wanted to ensure that \typetracer captured type for
  unevaluated arguments.  Not every argument of a function needs to be used
  during every function execution, and the approach we discussed above
  relies on argument promises being forced.  There are a few reasons for this.

One possibility is that the argument was simply unused given the control
flow of this particular execution (e.g. if the argument is only needed in
one branch of a conditional expression).  This mainly occurs when arguments have default
values for some parameters that are not critical to every execution of the
function.  In these cases, we still want to collect the type of the default
value.

Another possibility is that a value is used {\it without being forced}, as
is the case when values are ``metaprogrammed''.
This is common practice in e.g. the \code{tidyverse} ecosystem, as they use the metaprogramming capabilities of R to implement a DSL for data analysis.

\item Second, we wanted to account for formal arguments (arguments which are
  named in the function definition) that were {\it not} passed to the
  function. This represents cases where a function was not called with all of its
  arguments, and no default value was specified for missing arguments. 
  We record a ``missing'' type for such argument.

There are two obvious ways to deal with missing arguments: type them as bot
or as top.  As we are performing a dynamic analysis based on package test
code, we conservatively type these arguments as top, or \code{any}, as It is
impossible to know dynamically what all possible function behaviours are.
We discuss the prevalence of such any-typed arguments in Section~\ref{sec:evaluation}.

\item Third, we had to contend with R's C implementation sometimes
  performing longjumps, which discards current R calls from the call stack.
  In order to ensure that call traces are not lost when a longjump occurs,
  \typetracer intercepts the unwinding process initiated by the longjump and
  mimics the functionality of the functions being exited.  The R-dyntrace
  framework supplies the return value during the unwinding process (if there
  is indeed such a value), so we can capture it, analyze it, and include it
  in the call trace.  In certain situations, however, no return value can be
  obtained (e.g. a longjump occurred as part of some error handling), and
  then we simply record a ``jumped'' return value, which we can deal with in
  post-processing.  
  % These longjumps occur quite often, particularly when S3 dispatch occurs, as the R implementation optimizes returns from dispatched functions by initiating a longjump.

We primarily rely on 8 R-dyntrace callbacks: \texttt{closure\_entry},
\texttt{closure\_exit}, \texttt{builtin\_entry}, \texttt{builtin\_exit},
\texttt{special\_entry},\texttt{special\_exit},
\texttt{promise\_force\_entry}, and \texttt{promise\_force\_exit}.  These
callbacks fire when closures, builtin, and special functions are entered and
exited, and before and after the forcing of promises.  The function-related
callbacks are used mainly for bookkeeping: the analysis is notified that a
closure, builtin, or special has been entered/exited by pushing/popping the
call onto a stack.  The calls themselves store a {\it call trace} object,
where we store the type information that we collect, populated with initial
guesses of argument types (as discussed in
Section~\ref{subsec:typetracer-challenges}).  Further, the type of dispatch
is recorded on the {\tt \_entry} variants: recall that R can perform single or multiple dispatch on function arguments depending on their class.
We collect dispatch information in order to analyze dispatch usage patterns.

Once types have been simplified in this way, for each function we move to
consolidate its multiple call traces into a single function type.  We do
this by combining the observed types of each argument into one type according to subtyping rules that are part of our
design.  The precise subtyping rules we used will be discussed further in
Section~\ref{sec:typesystemdesign}.
If we cannot consolidate multiple observed argument types into a single type, we simply union the disparate types together.

While our analysis collects more detailed information than necessary for type generation, the information proved useful for building up a usage profile of the language. 
This profile informed our language design: based on the rich type information we collected, we were able to determine types that would be useful for capturing the observed behaviour.
We discuss this iterative, data-driven language design process in Section~\ref{types}, highlighting the reasoning behind our type system design.

%
%
%
%
\subsection{contractr}
\label{sec:contractr}

@@

@@

\AT{Include a figure? Histogram(s) of types?}

In order to count the occurrences of each type, we subdivided them into {\it kinds}: 
For our purposes, the {\it kind} of a type is equivalent to it's top-level type constructor, so all classes have kind ``class'', all lists have kind ``list'', and so on.
We found that classes are the most common kind of type, accounting for roughly 31\% of types.
\AT{Make sure to transform class<function> into any=>any.}
In order, the most common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%), factors (2\%), and tibbles (2\%).
Roughly 25\% of classes are part of R's base language implementation, the others being defined as part of some package's functionality.

Besides classes, scalars and vectors are the next most common kind, making up 41\% of types. with scalars making up 28.1\% of types and vectors 12.65\%.
Nulls and lists follow at 8\% and 7\% respectively, and the vararg type \ldots makes up 6.5\% of arguments.
This all totals up to over 90\% of types.
We will discuss the any type in more detail in Section~\ref{subsec:any}.

It would appear that our simple type annotations nicely account for the landscape of values in R.
\AT{Not sure what to say about classes.}
% Classes are a common kind of type in R, and for the purposes of this type annotation language allowing users to specify class names as part of class types is welcome: we saw that the breakdown of the class kind does not reveal many outlier classes,

@@

The polymorphism in R is best described by {\it ad hoc} polymorphism, wherein polymorphism is implemented by function and operator overloading (see S3 dispatch) rather than in some more fundamental way in the language design.
For the purposes of this evaluation, we will say that an argument is polymorphic if its type is a union of two or more types, in other words the argument has been observed to take more than one type of value.

\AT{We need better numbers on this.}
We found that 15\% of argument positions were polymorphic, and that X\% of functions were polymorphic.

Our type annotation language does not support sophisticated class typing, and much of the polymorphism in R is due to these classes, namely through the S3 dispatch mechanism. 
In future work, we aim to more fully handle S3 and objects in R.
For now, if we count out

@@

%
%
\subsubsection{Use of \code{any}}
\label{subsec:any}
